---
title: "Model exploration of AT–CTI Disequilibrium"
author: ""
format:
  html:
    code-overflow: wrap
    code-fold: true
    code-line-numbers: true
    code-tools:
      source: false
    toc: true
    toc-depth: 3
    theme:
      light: flatly
    number-sections: true
embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(patchwork)
library(broom)
library(knitr)
library(nimble)

theme_set(theme_classic(base_size = 12))

source(here::here("scripts/plot-cases.R"))
source(here::here("scripts/data-simulation.R"))
source(here::here("scripts/model-bayes.R"))
```

# Model formulation

## Motivation: the analogue of the Lotka--Volterra model

The classical Lotka--Volterra predator--prey model describes reciprocal interactions: prey abundance affects predator growth, while predators affect prey loss. Here, the ecological question is similar but different. Annual temperature ($x$) may influence the community temperature index ($\mu$), but $\mu$ does not feed back on annual temperature.

The causal direction is therefore

$$
x_t \longrightarrow \mu_{t+1},
$$

with no reciprocal term

$$
\mu_t \not\longrightarrow x_{t+1}.
$$

It is a one-way, discrete-time dynamic system that borrows the idea of coupling two state variables while allowing only one directional interaction.

## Discrete-time model

Let $x_t$ and $\mu_t$ denote annual temperature and the community temperature index at time $t$. Ignoring the error terms temporarily, define

$$
x_{t+1}=x_t+ax_t=(1+a)x_t,
$$

and

$$
\mu_{t+1}=\mu_t+b\mu_t+cx_t
          =(1+b)\mu_t+cx_t.
$$

In matrix form,

$$
\begin{pmatrix}
x_{t+1}\\
\mu_{t+1}
\end{pmatrix}
=
\begin{pmatrix}
1+a & 0\\
c & 1+b
\end{pmatrix}
\begin{pmatrix}
x_t\\
\mu_t
\end{pmatrix}.
$$

The zero in the upper-right corner encodes the one-way assumption: CTI does not affect subsequent AT.

## Interpretation of the three parameters

### Parameter $a$: temporal change in AT

$$
x_{t+1}=(1+a)x_t.
$$

- $a>0$: AT increases multiplicatively through time.
- $a=0$: AT remains constant.
- $a<0$: AT decreases multiplicatively.

For example, $a=0.02$ means that AT is multiplied by $1.02$ at each time step.

### Parameter $b$: internal persistence or relaxation of CTI

Ignoring AT temporarily,

$$
\mu_{t+1}=(1+b)\mu_t.
$$

Thus, $1+b$ is the autoregressive persistence of $\mu$.

- $-1<b<0$: CTI retains part of its previous state but relaxes through time.
- $b=0$: CTI is fully persistent in the absence of AT forcing.
- $b>0$: the internally propagated CTI component grows.

For example, $b=-0.20$ gives $1+b=0.80$, so 80% of the previous CTI state is retained in the internal component.

### Parameter $c$: one-way AT forcing on CTI

$$
\mu_{t+1}=(1+b)\mu_t+cx_t.
$$

The parameter $c$ measures the conditional effect of current AT on subsequent CTI after accounting for current CTI.

- $c>0$: higher AT predicts higher subsequent CTI.
- $c=0$: AT has no one-step effect on CTI.
- $c<0$: higher AT predicts lower subsequent CTI.

### Recovering the disequilibrium effect algebraically

Define climatic disequilibrium as

$$
D_t=x_t-\mu_t.
$$

Then

$$
x_t=\mu_t+D_t.
$$

Substituting this identity into the CTI equation gives

$$
\begin{aligned}
\mu_{t+1}
&=(1+b)\mu_t+cx_t\\
&=(1+b)\mu_t+c(\mu_t+D_t)\\
&=(1+b+c)\mu_t+cD_t.
\end{aligned}
$$

Therefore, we can do an algebraic reparameterization:

$$
\boxed{\mu_{t+1}=(1+b+c)\mu_t+cD_t.}
$$

### Assumption of an invariant $x=\mu$

The condition $c=a-b$ is derived by requiring the equality $x=\mu$ to remain unchanged over time.

Suppose that, at time $t$,

$$
x_t=\mu_t.
$$

According to the model,

$$
x_{t+1}=(1+a)x_t=(1+a)x,
$$

whereas

$$
\mu_{t+1}
=(1+b)\mu_t+cx_t
=(1+b+c)x.
$$

For the two variables to remain equal at the next time step, we require

$$
x_{t+1}=\mu_{t+1}.
$$

Therefore,

$$
(1+a)x=(1+b+c)x.
$$

For ($x\neq0$), this gives

$$
1+a=1+b+c,
$$

and hence

$$
\boxed{c=a-b.}
$$

Thus, $c=a-b$ is the condition under which $x_t=\mu_t$ implies $x_{t+1}=\mu_{t+1}$. In other words, the line $\mu=x$, or equivalently $D=x-\mu=0$, is an invariant line of the dynamic system..

# Simulation cases

We simulate four cases to illustrate the dynamics of the one-way AT--CTI model. The four cases are:

1. Constant AT; CTI starts below AT;
2. Constant AT; CTI starts above AT;
3. Increasing AT; CTI starts below AT;
4. Increasing AT; CTI starts above AT

For all four cases, `c` is omitted. The function therefore sets

$$
c=a-b.
$$

With $b=-0.20$, the disequilibrium dynamics become

$$
D_{t+1}=0.8D_t,
$$

so 80% of the previous disequilibrium is retained and 20% is removed at each time step.

```{r}
#| label: four-simulations

sim_cases <- list(
  generate_plot_at_cti(
    n_step = 30,
    AT0 = 20,
    CTI0 = 10,
    a = 0,
    b = -0.20,
    case = "1. Constant AT; CTI starts below AT"
  ),
  generate_plot_at_cti(
    n_step = 30,
    AT0 = 20,
    CTI0 = 30,
    a = 0,
    b = -0.20,
    case = "2. Constant AT; CTI starts above AT"
  ),
  generate_plot_at_cti(
    n_step = 30,
    AT0 = 20,
    CTI0 = 10,
    a = 0.02,
    b = -0.20,
    case = "3. Increasing AT; CTI starts below AT"
  ),
  generate_plot_at_cti(
    n_step = 30,
    AT0 = 20,
    CTI0 = 30,
    a = 0.02,
    b = -0.20,
    case = "4. Increasing AT; CTI starts above AT"
  )
)

sim_all <- bind_rows(sim_cases)
```

## AT--CTI state space

```{r}
#| label: figure-state-space
#| fig-width: 10
#| fig-height: 8

plots_all <- plot_at_cti_dynamics(sim_all, facet = TRUE)
plots_all$space
```

Interpretation:

- The dashed 1:1 line represents $D=0$.
- Cases 1 and 3 begin below the line because $x_0>\mu_0$, so $D_0>0$.
- Cases 2 and 4 begin above the line because $x_0<\mu_0$, so $D_0<0$.
- Under $c=a-b$ and $-1<b<0$, all trajectories approach the 1:1 line without overshooting.
- When $a=0.02$, the target itself moves because AT increases through time.

## First-difference AT--CTI state space

```{r}
#| label: figure-delta-space
#| fig-width: 10
#| fig-height: 8

plots_all$delta_space
```

Here, each point is $(\Delta x_t, \Delta\mu_t)$, where

$$
\Delta x_t=x_t-x_{t-1},\qquad
\Delta\mu_t=\mu_t-\mu_{t-1}.
$$

## Time--value trajectories

```{r}
#| label: figure-time-value
#| fig-width: 10
#| fig-height: 8

plots_all$time_value
```

## Disequilibrium through time

```{r}
#| label: figure-disequilibrium
#| fig-width: 10
#| fig-height: 7

sim_all |>
  ggplot(aes(x = time, y = D)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.7) +
  facet_wrap(~case, scales = "free_y") +
  labs(
    x = "Time",
    y = expression(D[t] == x[t] - mu[t]),
    title = "Disequilibrium dynamics"
  )
```

# Estimating the parameters from observations

## Model fitting

For empirical data, we can fit the one-way AT--CTI model using linear regression. The two equations are:

$$
x_{i,t+1}=\alpha_{x}+\phi_{x}x_{i,t}+\varepsilon^{x}_{i,t},
$$

$$
\mu_{i,t+1}=\alpha_{\mu}+\phi_{\mu}\mu_{i,t}+cx_{i,t}
+\varepsilon^{\mu}_{i,t}.
$$

The dynamic coefficients are still summarized as

$$
a=\phi_{x}-1,
\qquad
b=\phi_{\mu}-1.
$$

## What should be tested?

Three related but distinct questions can be evaluated.

### Directional AT or disequilibrium effect

$$
H_1:c>0.
$$

A positive $c$ means that, after accounting for current CTI, a more positive $D_t=x_t-\mu_t$ predicts a higher subsequent CTI.

### Is $\mu=x$ the invariant target line?

Define

$$
q=c-(a-b).
$$

Test

$$
H_0:q=0.
$$

Failure to reject $q=0$ is consistent with $D=0$ being invariant. It does not by itself prove convergence.

Because $a$ is estimated from the AT equation while $b$ and $c$ are estimated from the CTI equation, uncertainty should account for both equations. We could consider using a site-level bootstrap,  method to construct the confidence interval for $q$.

### Does disequilibrium converge?

When $q=0$,

$$
D_{t+1}=(1+b)D_t.
$$

Convergence requires

$$
|1+b|<1.
$$

Monotonic convergence without overshooting requires

$$
-1<b<0.
$$


# Simulations

## Generating simulation data

We first generate simulation data for multiple plots with shared dynamics,
different initial states, and plot-level random intercepts.

```{r generate-simulation}
sim_stochastic_plots <- simulate_at_cti_stochastic_plots(
  n_plot = 20,
  n_step = 15,
  AT0_min = 15,
  AT0_max = 20,
  CTI0_min = 10,
  CTI0_max = 25,
  a = 0.02,
  b = -0.20,
  sd_AT = 1,
  sd_CTI = 1,
  sigma_plot_AT = 0.5,
  sigma_plot_CTI = 0.5,
  case = "Stochastic plots: shared dynamics; different initial states",
  seed = 123
)
```

The same plotting function can be applied to the stochastic multi-plot data.
Each plot is shown as a trajectory.

```{r stochastic-plots}
#| fig-width: 7
#| fig-height: 7
stochastic_plots <- plot_at_cti_dynamics(
  sim_stochastic_plots |>
    mutate(case = paste("Plot", plot)),
  facet = F
)

stochastic_plots$space
```


## Using two independent models to estimate parameters

We can fit the AT and CTI equations independently using linear mixed-effects
models, with plot as a random intercept. The fixed effects are shared across
plots, while each plot has its own intercept.

The simulation includes plot-specific dynamic intercepts, so the random
intercept terms in the fitted models correspond to effects generated in the
data.

```{r freq-model}
sim_for_fit <- sim_stochastic_plots |>
  group_by(plot) |>
  arrange(time, .by_group = TRUE) |>
  mutate(
    AT_next = lead(AT),
    CTI_next = lead(CTI)
  ) |>
  ungroup() |>
  filter(!is.na(AT_next), !is.na(CTI_next))

# Fit the AT equation:
# AT_(t+1) = 1 + a + a_plot + error
fit_AT <- lme4::lmer(
  AT_next ~ AT + (1 | plot),
  data = sim_for_fit,
  REML = FALSE
)

# Fit the CTI equation:
# CTI_(t+1) = 1 + b + c + b_plot + error
fit_CTI <- lme4::lmer(
  CTI_next ~ CTI + AT + (1 | plot),
  data = sim_for_fit,
  REML = FALSE
)

z_critical <- qnorm(0.975)

AT_fixed <- lme4::fixef(fit_AT)["AT"]
AT_vcov <- as.matrix(vcov(fit_AT))
AT_se <- sqrt(AT_vcov["AT", "AT"])

CTI_fixed <- lme4::fixef(fit_CTI)[c("CTI", "AT")]
CTI_vcov <- as.matrix(vcov(fit_CTI))
CTI_se <- sqrt(diag(CTI_vcov)[c("CTI", "AT")])

parameter_estimates <- tibble(
  parameter = c("a", "b", "c"),
  estimate = c(
    AT_fixed - 1,
    CTI_fixed["CTI"] - 1,
    CTI_fixed["AT"]
  ),
  std_error = c(
    AT_se,
    CTI_se["CTI"],
    CTI_se["AT"]
  )
) |>
  mutate(
    conf_low = estimate - z_critical * std_error,
    conf_high = estimate + z_critical * std_error
  )
parameter_estimates
```

## Using bayesian methods

We can estimate the shared dynamic parameters jointly with the plot-level
random effects. The model uses the same two transition equations as above. The parameters
`a`, `b`, and `c` are shared across plots, while the two equations have
separate plot-level random intercepts and residual standard deviations.

```{r}
#| message: false
#| warning: false
#| label: posterior-distributions
#| fig-width: 10
#| fig-height: 5

bayes_fit <- fit_at_cti_bayes(
  sim = sim_stochastic_plots,
  niter = 15000,
  nburnin = 5000,
  thin = 4,
  nchains = 10
)

bayes_fit$posterior_summary
bayes_fit$posterior_plot
```

